# Logical Operators {#sec-logical}

Logical operators are able to check whether certain conditions are `TRUE` or `FALSE`. The standard logical operators remain fairly consistent with other programs for operators such as `==`, `!=`, `<`, `<=`, `>`, `>=`; however, the `And` and `Or` operators are provided by `&` and `|` respectively. This could differ from previously learnt languages. `&&` and `||` are also logical operators available.

As you might expect, there are subtle differences between these latter operators. To see the difference, consider `&` and `&&`.

```{r}

a <- c(1,2,5)
b <- a

HecticCalculation <- function(){
  
  print('I am doing a hectic calculation, I promise')
  
  return(T)
}

# Operator comparison

if(all(a != b) & HecticCalculation()){
  print('I should be unreachable, first condition false')
}
```

```{r}

if(all(a != b) && HecticCalculation()){
  print('I should be unreachable, first condition false')
}

```

```{r}
if(all(a == b) && HecticCalculation()){
  print('I can run.')
}

```

The first `if` statement runs the *hectic calculation* because it checks **both** of the conditions on either side of it!

The `&&` does **not** run the hectic calculation for the second `if` statement because the first condition fails (it prints nothing). For final `if` statement runs the *hectic calculation*, since the first condition passed, therefore the second condition is checked.

This type of logical behaviour can be very useful when dealing with difficult computations if used effectively, avoiding unnecessary function calls. The same can be achieved with `||`.

In the above code, we have used the `if` statement which runs code *if* a condition is met. With this, we have the `else` and `if else` statements. They generally follow a structure like this:

```{r}
aCondition = F
anotherCondition = T

if (aCondition){
  "Check original condition"
}else if (anotherCondition){
  "Check another condition IF `aCondition` is false"
}else {
  "If all of the above are false, then do this"
}
```
