# Loops {#sec-day2-loops}

There are two common types of loops in programming:

1.  *For* loops
2.  *While* loops

These are consistent with other programming languages. Below are some examples to illustrate the required *R* syntax.

## For Loops

The general structure of the syntax is:

`for (indicator in iterable){ ... }`

where the indicator is a variable which iterates through an object which is *iterable*. Therefore, indicator can be anything.

```{r}

a <- 1:10

sum = 0 
for (i in a){
  sum = sum + i
}
print(sum)

sentence <- "hello world and some other words"
anIterable <- strsplit(sentence, " ")
for (word in anIterable){
  print(word)
}

```

We can even have nested for loops, which is when we have a for loop within another for loop!

## While Loops

The general structure of the syntax is:

`While (condition) { ... }`

where the condition is a Boolean (`TRUE` or `FALSE`). It could be the result of many conditions combined with logical operators. This loop will continue until this condition is false.

```{r}
max = 10
x <- 0; i <- 0
while(x <= max){
  i <- i + 1
  x <- x + 2 * i 
}
x
```
